---
title: "Filter scRNA-Seq and VDJ data"
output: html_document
date: "2024-07-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# START
* Workflow for processing combined analysis object
* Prior to running this code set working directory to 'impact-analysis/code'

```{r,warning=F,message=F}
# Set Working Directory before starting

# Immcantation
suppressPackageStartupMessages(library(alakazam))
suppressPackageStartupMessages(library(shazam))
suppressPackageStartupMessages(library(tigger))
suppressPackageStartupMessages(library(dowser))
suppressPackageStartupMessages(library(airr))

suppressPackageStartupMessages(library(rjson))
suppressPackageStartupMessages(library(glmGamPoi))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(dittoSeq))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(devtools))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(impactSingleCellToolkit))

# Increase size of ENV
options(future.globals.maxSize= 891289600)
set.seed(12345)
```


# 1. Load inputs

### a. params
```{r}
# INPUT and OUTPUT Directories
param_file_fh = "../input/input_one_patient_analysis.json"
params        = fromJSON(file = param_file_fh)

# INPUT
metadata_fh   = params$metadata_file
metadata      = read.csv(metadata_fh,stringsAsFactors = F,check.names = F)
THREADS       = params$threads

# subset metadata
metadata_gex  <- metadata[metadata$type %in% c("counts_gex"),]
metadata_bcr  <- metadata[metadata$type %in% c("vdj_b"),]
metadata_tcr  <- metadata[metadata$type %in% c("vdj_t"),]

# OUTPUT Results Directories
plot_dir     = params$plot_directory
objects_dir  = params$objects_directory
if(!file.exists(plot_dir)){dir.create(plot_dir,recursive = TRUE)}
if(!file.exists(objects_dir)){dir.create(objects_dir,recursive = TRUE)}
```

### b. processed GEX/CSP object
```{r}
fh_processed_seurat_obj <- file.path(objects_dir,"seurat_obj.processed.rds")
rnaseq_obj              <- readRDS(fh_processed_seurat_obj)
```


# 2. Load and format TCR/BCR repertoire results
* Clonotype definitions for BCR are based on the Heavy chain only
  * Plots CDR3 distance distribution among BCRs and come up with a new threshold
* Clonotype definitions for TCR include both Alpha and Beta chains

### a. Load Immcantation Results
```{r}
# File paths
bcr_heavy_fh <- "../../Results/results_immcantation_one_patient/BCR_CSFPB/BCR_CSFPB_heavy_germ-pass.tsv"
bcr_light_fh <- "../../Results/results_immcantation_one_patient/BCR_CSFPB/BCR_CSFPB_light_germ-pass.tsv"
tcr_beta_fh  <- "../../Results/results_immcantation_one_patient/TCR_CSFPB/TCR_CSFPB_heavy_germ-pass.tsv"
tcr_alpha_fh <- "../../Results/results_immcantation_one_patient/TCR_CSFPB/TCR_CSFPB_light_germ-pass.tsv"

# TCR
db.tra   <- read_rearrangement(tcr_alpha_fh) %>% as.data.frame()
db.trb   <- read_rearrangement(tcr_beta_fh) %>% as.data.frame()
# BCR
db.heavy <- read_rearrangement(bcr_heavy_fh) %>% as.data.frame()
db.light <- read_rearrangement(bcr_light_fh) %>% as.data.frame()
```

### b. filter contigs and merge all chains
* All contigs are in frame
```{r}
bcr_df <- mergeVDJresults(db.heavy,db.light,umi.thresh = 3,assay = "bcr")
tcr_df <- mergeVDJresults(db.trb,db.tra,umi.thresh = 2, assay = "tcr")

# format VDJ unique cell IDs to match scRNA-Seq (function only for this workshop)
formatUniqueIDcolum <- function(uni_id_column,sep = ":") {
  samples <- tstrsplit(uni_id_column, sep)[[1]]
  cells   <- tstrsplit(uni_id_column, sep)[[2]]
  fmt_ids <- paste(cells,samples,sep = "-")
  
  return(fmt_ids)
}

bcr_df$unique_cell_id <- formatUniqueIDcolum(bcr_df$unique_cell_id)
tcr_df$unique_cell_id <- formatUniqueIDcolum(tcr_df$unique_cell_id)
```


# 3. Only analyze cells shared across all datasets

### a. get vector of overlapping cells
```{r}
cells_rnaseq      <- colnames(rnaseq_obj)
cells_vdj         <- unique(c(bcr_df$unique_cell_id,
                              tcr_df$unique_cell_id))
overlapping_cells <- cells_rnaseq[cells_rnaseq %in% cells_vdj]
```


### b. filter rna-seq and vdj data
* Even after filtering for cells that overlap with BCR/TCR data, there are still sparse clusters of cells annotated as different cell types
```{r,fig.height=6,fig.width=14}
# generate UMAP of cell type annotations pre-filtering
p1_pre_filter <- DimPlot(object = rnaseq_obj, group.by="predicted.celltype.l1", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")

# only keep cells with vdj
rnaseq_obj     <- rnaseq_obj[,colnames(rnaseq_obj) %in% overlapping_cells]
bcr_df         <- bcr_df[bcr_df$unique_cell_id %in% overlapping_cells,]
tcr_df         <- tcr_df[tcr_df$unique_cell_id %in% overlapping_cells,]

# post filter umap
p2_post_filter <- DimPlot(object = rnaseq_obj, group.by="predicted.celltype.l1", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")

plot_grid(p1_pre_filter,p2_post_filter,ncol = 2)
```


# 4. Add metadata columns to GEX and VDJ data

### a. rnaseq object
```{r}
createMetadataColumn <- function(sample_vec, metadata,sample.col = "sample",meta.col = "SampleType") {
  new_col <- sample_vec
  uni_labels <- metadata[,meta.col] %>% unique()
  
  for (categ in uni_labels) {
    categ_samples <- metadata[metadata[,meta.col] %in% categ,sample.col] %>% unique()
    new_col[new_col %in% categ_samples] <- categ
  }
  
  return(new_col)
}

rnaseq_obj@meta.data$compartment <- createMetadataColumn(rnaseq_obj@meta.data$sample,metadata = metadata_gex,meta.col = "SampleType" )
rnaseq_obj@meta.data$status <- createMetadataColumn(rnaseq_obj@meta.data$sample,metadata = metadata_gex,meta.col = "Project" )
```

### b. VDJ
```{r}
# Add sample column
tcr_df$sample <- tstrsplit(tcr_df$unique_cell_id,"-")[[2]]
bcr_df$sample <- tstrsplit(bcr_df$unique_cell_id,"-")[[2]]

# Add column for Compartment
tcr_df$COMPARTMENT <- createMetadataColumn(tcr_df$sample,metadata = metadata_tcr,meta.col = "SampleType" )
bcr_df$COMPARTMENT <- createMetadataColumn(bcr_df$sample,metadata = metadata_bcr,meta.col = "SampleType" )

# Sample status
tcr_df$STATUS <- createMetadataColumn(tcr_df$sample,metadata = metadata_tcr,meta.col = "Project" )
bcr_df$STATUS <- createMetadataColumn(bcr_df$sample,metadata = metadata_bcr,meta.col = "Project" )
```


# 5. Re-cluster


